Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
incremental edits -- still much to do

@@

- higher level presentation of the problem to the computer, as
  in R's `nls` versus packages `minpack.lm` and `nlsr`, or as
  a general optimization problem.

@@

### Issues with this structure

A major criticism of this approach is that the collection of information needed
at any one time -- the "environment" in which residuals and jacobians are calculated,
as well as the solution methods -- is not at all well-documented. If the software is
to be updated or maintained, the location and purpose of the many elements that
reside in different environments is difficult to determine without much effort.
In my opinion, this renders the structure prone to collapse if there are infrastructure
or other changes. Documentation would go a long way to correcting this problem, but
even better would be clearer structuring. Separation of the problem from the method
would also be useful. At the time of writing of this document, the Gauss-Newton 
METHOD is embedded in the functions created by `nlsModel()` for a given PROBLEM.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements $\sqrt {(\lambda)}$ times the diagonal stabilization elements
we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
of the Gauss-Newton.

@@

The "original" Gauss-Newton iteration offers no guidance in how to deal with a
situation where $x_{new}$ has a poorer fit than $x$. @Hartley61 and @Marquardt1963,
along with most other practical methods, need to recompute the (weighted) residuals
and consequent loss function (sum of squares or deviance) for each value of a step-length
or stabilization parameter ($\lambda$) until either a "better" point $x_{new}$ is found
or we terminate the search.

Thus it is useful to be able to separate the computation of residuals from that of the
Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
historical reasons. This is the setup for both the finite difference approximations of
`nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
as well as the `rjfun()` generated by `nlsr::model2rjfun()`. Nevertheless, it is likley worth 
avoiding duplication if possible, though the computation of Jacobian elements uses a lot of the
same steps as the computation of the residuals.

?? need to explore timings

Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
situations:

- saving residual values when computing derivatives via finite differences
- separation of the Jacobian routine completely from that of the residuals
- other measures?

@@

If the choice of approach to Gauss-Newton or Marquardt is to build the normal
equations and hence the sum of squares and cross products (SSCP) matrix, we
know by construnction that this is a symmetric matrix and also non-negative definite.
(Generally, unless we have a strict linear dependence between the columns of
the Jacobian, we can say "positive definite".)
If we use an SSCP form, we can apply algorithms that specifically take advantage of both these
properties, including programs that store the working matrix in special formats. 
Some of these are Algorithms 7, 8 and 9 of Nashlib (@jncnm79). Algorithms 7 and 8 are the
Cholesky decomposition and back-solution using a vector of length `n*(n+1)/2`
to store just the lower triangle of the SSCP matrix. Algorithm 9 inverts this
matrix *in situ*.

@@

@@

In @Nash1987a, each major iteration generates a $A = J^t J$ matrix and the right hand side $-g = J^t r$
at parameters $x$ using functions for the residuals and Jacobian.
The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements (row-wise ordering of $C$). 
The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ is added to each,
where $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in @jn77ima to
overcome potential underflow to zero of the accumulation of the diagonal elements in low-precision
arithmetic. Let us call the augmented SSCP matrix $C$.

Solving $C \delta = -g$, we compute the new sum of squares at $x + \delta$. If this value is greater than
or equal to the current best (i.e., lowest) sum of squares, we increase $\lambda$ and generate a new
$C$ matrix. Note that we do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
adjustment.

The code in @Nash1987a is much more complicated than this description suggests because of

@@

- tests for "non-computability", where we allow the user to provide an explicit flag that the function cannot
  be computed at the trial parameters. This was included to avoid some of the weaknesses in evaluation of 
  special functions, e.g., `log()` for small arguments or `exp()` for large ones, or for problem specific
  situations where some parameter combinations might be considered inadmissible.

@@

That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.

>>>>

<!-- expand ??, include some BHHH stuff -->

@@

- higher level presentation of the problem to the computer, as
  in R's `nls` versus packages `minpack.lm` and `nlsr`, or as
  a general optimization problem

- incorporation of constraints on the parameters, namely bounds,
  masks or inequalities.

@@

### Issues with the `nls()` structure

A major criticism of the `nls()` approach is that the collection of information needed
at any one time -- the "environment" in which residuals and jacobians are calculated,
as well as the solution methods -- is not sufficiently well-documented. If the software is
to be updated or maintained, the location and purpose of the many elements that
reside in different environments is difficult to determine without effort.
In my opinion, this renders the structure prone to collapse if there are infrastructure
or other changes. Documentation would go a long way to correcting this problem, but
even better would be clearer structuring. Separation of the problem from the method
would also be useful. At the time of writing of this document, the Gauss-Newton 
METHOD is embedded in the functions created by `nlsModel()` for a given PROBLEM.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements $\sqrt {(\lambda)}$ times the square roots of the 
diagonal stabilization elements
we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
of the Gauss-Newton.

@@

The "original" Gauss-Newton iteration offers no guidance in how to deal with a
situation where $x_{new}$ has a poorer fit than $x$. @Hartley61 and @Marquardt1963,
along with most other practical methods, need to recompute the (weighted) residuals
and consequent loss function (sum of squares or deviance) for each value of a step-length
or stabilization parameter ($\lambda$) until either a "better" point $x_{new}$ is found
or we terminate the search. However, as we have seen, Hartley does not have to adjust
the search direction.

Thus it could be helpful to separate the computation of residuals from that of the
Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
historical reasons. This is the setup for both the finite difference approximations of
`nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
as well as the `rjfun()` generated by `nlsr::model2rjfun()`. On the other hand, 
the computation of Jacobian elements can share many calculations with the computation of the residuals. 
Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
situations, but is likely of importance only to users with extremely time consuming problems.

@@

If the choice of approach to Gauss-Newton or Marquardt is to build the normal
equations and hence the sum of squares and cross products (SSCP) matrix, we
know by construction that matrix is symmetric and also non-negative definite.
(Generally, unless we have a strict linear dependence between the columns of
the Jacobian, we can say "positive definite".)
If we use an SSCP form, we can apply algorithms that specifically take advantage of both these
properties, including programs that store the working matrix in special formats. 
Some of these are Algorithms 7, 8 and 9 of Nashlib (@jncnm79). Algorithms 7 and 8 are the
Cholesky decomposition and back-solution using a vector of length `n*(n+1)/2`
to store just the lower triangle of the SSCP matrix. Algorithm 9 inverts this
matrix *in situ*.

@@

?? sort out $A$ and $C$

@@

In @Nash1987a, each major iteration generates a $A = J^t J$ matrix and the right hand side $-g = J^t r$
at parameters $x$ using functions for the residuals and Jacobian.
The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements.
The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ is added to each,
then saved in matrix $C$. $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in
@jn77ima to
overcome potential underflow to zero of the accumulation of the diagonal elements in low-precision
arithmetic.

Solving $C \delta = -g$, we compute the new sum of squares at $x + \delta$. If this value is greater than
or equal to the current best (i.e., lowest) sum of squares, we increase $\lambda$ and generate a new
$C$ matrix. We do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
adjustment to $A$ to get a new $C$.

The code in @Nash1987a is much more complicated than this description, since we should address

@@

- tests for "non-computability", where we allow the user to provide an explicit flag that the function cannot
  be computed at the trial parameters. This was included to avoid some of the weaknesses in evaluation of 
  special functions, e.g., `log()` for small arguments or `exp()` for large ones, or for problem specific
  situations where some parameter combinations might be considered inadmissible. (Readers may note that
  the codes were written before @IEEE754-1985 was applied to most BASICs. The author was a member of
  that standardization committee.)

@@

That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newtx
